---
title: "LEGO EDA"
author: "Ansgar Wolsing"
date: "2022-09-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggtext)
library(lubridate)
library(here)

# Most common themes
```

```{r read_data, cache=TRUE}
base_path <- here::here("2022", "36")

#' Source: https://rebrickable.com/downloads/
#' Download data from website and extract the archives into the data directory

data_path <- here(base_path, "data")
files <- list.files(data_path, pattern = ".*\\.csv")
lego <- map(file.path(data_path, files), read_csv) %>% 
  set_names(str_remove(files, "\\.csv"))

```

## Exploring the tables

### Most frequent themes

Sets are assigned to themes via key `theme_id`.

```{r}

lego$sets %>% 
  left_join(lego$themes, by = c("theme_id" = "id"), suffix = c("", ".theme")) %>% 
  count(name.theme, sort = TRUE)

```

Which sets are part of LEGO gear?

```{r}

lego$sets %>% 
  left_join(lego$themes, by = c("theme_id" = "id"), suffix = c("", ".theme")) %>%
  filter(name.theme == "Gear") %>% 
  select(name, year) %>% 
  head(100)

```
Looks like theme **"Gear" should be excluded** since it's rather merch and video games.


### Colors

#### Colors from a particular set

```{r}
set <- "42119-1"

lego$sets %>% 
  filter(set_num == set) %>% 
  # join inventories
  inner_join(lego$inventories, by = "set_num") %>% 
  # join with inventory and parts to get the color ids of the parts + the parts' quantity
  inner_join(lego$inventory_parts, by = c("id" = "inventory_id")) %>% 
  inner_join(lego$colors, by = c("color_id" = "id"), suffix = c("", ".color")) %>% 
  select(set_num, name, year, num_parts, quantity, color_id, name.color, rgb)
  
```
#### Now all sets, aggregate by set and color

```{r}

colors_in_sets <- lego$sets %>% 
  # join inventories
  inner_join(lego$inventories, by = "set_num") %>% 
  # join with inventory and parts to get the color ids of the parts + the parts' quantity
  inner_join(lego$inventory_parts, by = c("id" = "inventory_id")) %>% 
  inner_join(lego$colors, by = c("color_id" = "id"), suffix = c("", ".color")) %>% 
  # exclude spare parts
  filter(!is_spare) %>% 
  count(set_num, year, name.color, color_id, rgb, wt = quantity, name = "quantity") %>% 
  mutate(rgb = paste0("#", rgb))

colors_in_sets

colors_in_sets %>% 
  filter(set_num == "001-1")

```


## Custom chart theme

```{r default_theme}

custom_theme <- function() {
  theme_void(base_family = "Roboto Condensed") +
  theme(
    plot.background = element_rect(color = "grey95", fill = "grey95"),
    plot.margin = margin(6, 6, 6, 6),
    text = element_text(color = "grey30"),
    plot.title = element_markdown(face = "bold", color = "black", size = 16, hjust = 0.5),
    plot.subtitle = element_textbox(hjust = 0.5, halign = 0.5, lineheight = 1.1, width = 0.85, size = 10),
    plot.caption = element_markdown(hjust = 0, lineheight = 1.1),
    strip.text = element_text(face = "bold", size = 12, hjust = 0.5),
    axis.text = element_text(),
    axis.title = element_text(),
    axis.title.y = element_text(angle = 90)
  )
}

# Some tweets for the treemap
theme_treemap <- function(x) {
  custom_theme() +
  theme(
    axis.text = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
  )
}

theme_set(custom_theme())
```


Standard plot caption:
```{r}
plot_caption <- "**Source:** Rebrickable. **Visualization:** Ansgar Wolsing"
```



### Number of sets per year

```{r}
lego$set %>% 
 ggplot(aes(year)) +
  geom_bar()

```

Start at 1975, 1980 maybe?


**Are there duplicate sets due to multiple inventories per set?**


### Colors per release year

```{r}
colors_in_sets %>% 
  count(year, rgb, wt = quantity, name = "quantity") %>%
  filter(year >= 1980) %>% 
  ggplot(aes(year, quantity, fill = rgb)) +
  geom_col() +
  scale_fill_identity()

```

```{r}
colors_in_sets %>% 
  count(year, rgb, wt = quantity, name = "quantity") %>% 
  arrange(rgb) %>% 
  mutate(rgb = fct_inorder(rgb)) %>% 
  filter(year >= 1970) %>%
  ggplot(aes(year, quantity, fill = rgb)) +
  geom_col(position = "fill") +
  scale_fill_identity()

```


```{r}
colors_in_sets %>% 
  count(year, rgb, wt = quantity, name = "quantity") %>% 
  # start in 1970/1980, end with the second last year
  filter(year >= 1970, year < max(year)) %>% 
  ggplot(aes(year, quantity, fill = rgb)) +
  geom_area() +
  scale_fill_identity()

```


#### Streamgraph

```{r}
library(ggstream)

colors_in_sets %>% 
  count(year, rgb, wt = quantity, name = "quantity") %>% 
  mutate(rgb = fct_reorder(rgb, quantity)) %>% 
  filter(year >= 1970) %>%
  ggplot(aes(year, quantity, fill = rgb)) +
  geom_stream(type = "ridge") +
  scale_fill_identity()
```



```{r}

colors_in_sets %>% 
  count(year, rgb, wt = quantity, name = "quantity") %>% 
  mutate(rgb = fct_reorder(rgb, quantity)) %>% 
  filter(year >= 1970) %>%
  ggplot(aes(year, quantity, fill = rgb)) +
  geom_stream(type = "proportional") +
  scale_fill_identity()
```



## Treemap comparing years

### Prepare data for treemaps

```{r}
library(treemapify)

colors_year <- colors_in_sets %>% 
  filter(name.color != "[Unknown]") %>% 
  count(year, name.color, rgb, wt = quantity, name = "quantity")

colors_years_agg_1 <- filter(colors_year, year >= 1987 & year <= 1992) %>% 
    # summarize for the full period
    count(rgb, wt = quantity, name = "quantity")

# Create a dataset with treemap dimensions - years 1987 to 1992
treemap_df_1 <- treemapify(
  colors_years_agg_1,
  area = "quantity",
  subgroup = "rgb",
  layout = "squarified"
)


colors_years_agg_2 <- filter(colors_year, year >= 2017 & year <= 2022) %>% 
    # summarize for the full period
    count(rgb, wt = quantity, name = "quantity")
  
treemap_df_2 <- treemapify(
  colors_years_agg_2,
  area = "quantity",
  subgroup = "rgb",
  layout = "squarified"
)

```


### Plot the treemaps

Plotting function

```{r}
# Show labels only for the more common colors above the threshold
color_share_threshold <- 0.05

plot_color_treemap <- function(df, colors_agg, color_share_threshold = 0.05) {
  df %>% 
    inner_join(colors_agg, by = "rgb") %>% 
    mutate(
      color_share = quantity / sum(quantity),
      label = paste(rgb, scales::percent(color_share, accuracy = 0.1), sep = "<br>")) %>% 
    ggplot(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax)) +
    geom_rect(aes(fill = rgb), color = "white", size = 0.5) +
    geom_richtext(
      data = ~subset(., color_share > color_share_threshold),
      aes(x = xmin, y = ymax, label = label),
      inherit.aes = FALSE, hjust = 0, vjust = 1, fill = alpha("white", 0.5), label.size = 0,
      nudge_x = 0.01, nudge_y = -0.015,
      family = "Roboto Condensed", lineheight = 1, size = 3) +
    scale_fill_identity() +
    theme_treemap()

}

```


1987 to 1992

```{r}
plot_color_treemap(treemap_df_1, colors_years_agg_1)  
```


2017 to 2022
```{r}
plot_color_treemap(treemap_df_2, colors_years_agg_2, color_share_threshold = 0.03)  
```


## "Cluster" by colors

Add a subgroup for color variants so that the color treemap for today shows the greater diversity of color shades

### By name

```{r}
color_group_names <- c("Red", "Blue", "Green", "Yellow", "Gray", "White", "Black", 
                       "Brown", "Pink", "Violet", "Purple", "Orange")

purple <- "Pink/Violet/Purple"

color_groups <- lego$colors %>% 
  filter(name != "[Unknown]") %>% 
  transmute(name, 
            rgb = paste0("#", rgb),
            color_group = str_extract(name, paste(color_group_names, collapse = "|")),
            color_group = replace_na(color_group, "Other"),
            color_group = case_when(
              color_group %in% c("Pink", "Violet", "Purple") ~ purple,
              name %in% c("Magenta", "Lavender", "Coral") ~ purple,
              str_detect(name, "Lavender") ~ purple,
              str_detect(name, "(Lemon|Turquoise)") ~ "Green",
              name == "Copper" ~ "Brown",
              name == "Rust" ~ "Red",
              str_detect(name, "Aqua") ~ "Green",
              str_detect(name, "Lime") ~ "Green",
              str_detect(name, "Salmon") ~ purple,
              str_detect(name, "Azure") ~ "Blue",
              str_detect(name, "\\b(Tan|Gold|Nougat)") ~ "Brown",
              name == "Chrome Antique Brass" ~ "Brown",
              TRUE ~ color_group
            ))
color_groups

count(color_groups, color_group, sort = TRUE)

```

```{r}

colors_years_agg_color_group_1 <- filter(colors_year, year >= 1987 & year <= 1992) %>% 
    # summarize for the full period
    count(rgb, wt = quantity, name = "quantity") %>% 
  inner_join(color_groups, by = "rgb") %>% 
  group_by(rgb) %>% 
  slice_head(n = 1) %>% 
  ungroup()
  
treemap_df_color_group_1 <- treemapify(
  colors_years_agg_color_group_1,
  area = "quantity",
  subgroup = "color_group",
  subgroup2 = "rgb",
  layout = "squarified"
)

colors_years_agg_color_group_2 <- filter(colors_year, year >= 2017 & year <= 2022) %>% 
    # summarize for the full period
    count(rgb, wt = quantity, name = "quantity") %>% 
  inner_join(color_groups, by = "rgb") %>% 
  group_by(rgb) %>% 
  slice_head(n = 1) %>% 
  ungroup()
  
treemap_df_color_group_2 <- treemapify(
  colors_years_agg_color_group_2,
  area = "quantity",
  subgroup = "color_group",
  subgroup2 = "rgb",
  layout = "squarified"
)
```


```{r}

plot_color_treemap_grouped <- function(df, color_agg) {
   df %>%  #treemap_df_color_group_2 
  inner_join(color_agg, by = c("rgb", "name", "color_group")) %>% #colors_years_agg_color_group_2
  mutate(
    color_share = quantity / sum(quantity),
    label = paste(rgb, scales::percent(color_share, accuracy = 0.1), sep = "<br>")) %>% 
  # calculate coordinates for subgroup borders manually
  group_by(color_group) %>% 
  mutate(
    subgroup_xmin = min(xmin),
    subgroup_xmax = max(xmax),
    subgroup_ymin = min(ymin),
    subgroup_ymax = max(ymax)
  ) %>% 
  ungroup %>% 
  ggplot(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax)) +
  geom_rect(aes(fill = rgb), color = "white", size = 0.1) +
  geom_rect(
    aes(xmin = subgroup_xmin, xmax = subgroup_xmax,
        ymin = subgroup_ymin, ymax = subgroup_ymax),
    color = "white", fill = NA, size = 0.8) +
  # # debug: print color names
  # ggrepel::geom_text_repel(aes(x = xmin, y = ymax, label = name), size = 2) +
  geom_richtext(
    data = ~subset(., color_share > color_share_threshold),
    aes(x = xmin, y = ymax, label = label),
    inherit.aes = FALSE, hjust = 0, vjust = 1, fill = alpha("white", 0.5), label.size = 0,
    nudge_x = 0.01, nudge_y = -0.015,
    family = "Roboto Condensed", lineheight = 1, size = 3) +
  scale_fill_identity() +
  theme_treemap()
}

```


```{r}
p_tm_1 <- plot_color_treemap_grouped(treemap_df_color_group_1, colors_years_agg_color_group_1)
p_tm_1
```

```{r}
p_tm_2 <- plot_color_treemap_grouped(treemap_df_color_group_2, colors_years_agg_color_group_2)
p_tm_2
```

### Combine both periods in one plot

```{r}
library(patchwork)

p_tm_1 + labs(subtitle = "Sets released<br>**1987 to 1992**") + 
  p_tm_2 + labs(subtitle = "Sets released<br>**2017 to 2022**") + 
  plot_annotation(
    title = "The Diversified Colors of LEGO\U2122",
    subtitle = "The number of different colors has changed substantially over the last decades.
    Each treemap presents the share of parts with the respective colors.<br>
    The number of sets released has increased as well, which is not shown in this chart.",
    caption = plot_caption
  ) +
  plot_layout() 

ggsave(here(base_path, "plots", "lego-colors-treemap-1992-vs-2022.png"),
       width = 10, height = 6)

```


## Most frequent colors share of all parts

Which were the most common colors in 1987 to 1992?

```{r}
frequent_colors_1 <- colors_years_agg_color_group_1 %>% 
  arrange(-quantity) %>% 
  mutate(share = quantity / sum(quantity)) %>% 
  filter(share > 0.05)

frequent_colors_1
main_colors <- frequent_colors_1$rgb
sum(frequent_colors_1$share)

```

What's their share in 2017 to 2022?

```{r}
colors_years_agg_color_group_2 %>% 
  arrange(-quantity) %>% 
  mutate(share = quantity / sum(quantity)) %>% 
  filter(rgb %in% main_colors)

colors_years_agg_color_group_2 %>% 
  arrange(-quantity) %>% 
  mutate(share = quantity / sum(quantity)) %>% 
  filter(rgb %in% main_colors) %>% 
  summarize(combined_share = sum(share))


```

### Year-by-year trend


The default gray from 1987 (Light Gray, #C91A09) seems to have been replaced with another one in the early 2000s (Light Bluish Gray, #A0A5A9).

```{r}
# recode gray tones
old_light_gray <- "#C91A09"
new_light_bluish_gray <- "#A0A5A9"
main_colors <- c(main_colors, new_light_bluish_gray)

light_gray_note <- "The default gray from 1987 (Light Gray, #C91A09) has been replaced
      with another one in the early 2000s (Light Bluish Gray, #A0A5A9).<br>This is accounted for in the plot.<br>"

colors_year %>% 
  filter(year >= 1987) %>% 
  group_by(year) %>% 
  mutate(share = quantity / sum(quantity)) %>% 
  ungroup() %>% 
  filter(rgb %in% main_colors) %>% 
  group_by(year, rgb) %>% 
  summarize(
    quantity = sum(quantity),
    share = sum(share)
  ) %>% 
  # calculate the combined share of the main colors
  summarize(combined_share = sum(share)) %>% 
  ggplot(aes(year, combined_share)) +
  geom_line(size = 1) +
  geom_point(
    data = ~subset(., year == min(year) | year == max(year)),
    size = 2
  ) +
  geom_text(
    data = ~subset(., year == min(year) | year == max(year)),
    aes(label = scales::percent(combined_share), hjust = ifelse(year == min(year), 1.3, -0.3)),
    size = 4
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0.25, NA)) +
  coord_cartesian(clip = "off") +
  labs(
    title = "In 1987, 6 colors accounted for almost all LEGO parts",
    subtitle = "Combined share of the 6 main colors from 1987",
    caption = paste(
      light_gray_note, 
      plot_caption, sep = "<br>"),
    x = NULL, y = NULL
  )

ggsave(here(base_path, "plots", "lego-colors-main-colors-line.png"), width = 8, height = 6)
```

### Area chart


```{r}

colors_year %>% 
  filter(year >= 1987) %>% 
  group_by(year) %>% 
  mutate(share = quantity / sum(quantity)) %>% 
  ungroup() %>% 
  filter(rgb %in% main_colors) %>% 
  group_by(year, rgb) %>% 
  summarize(
    quantity = sum(quantity),
    share = sum(share),
    .groups = "drop"
  ) %>% 
  mutate(rgb = fct_reorder(rgb, share)) %>% 
  ggplot(aes(year, share, fill = rgb)) +
  geom_area() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  scale_fill_identity() +
  coord_cartesian(clip = "off", expand = FALSE) +
  labs(
    title = "In 1987, 6 colors accounted for almost all LEGO parts",
    subtitle = "Combined share of the 6 main colors from 1987",
    caption = paste(
      light_gray_note, 
      plot_caption, sep = "<br>"),
    x = "Release year", y = NULL
  ) +
  theme(
    panel.background = element_rect(color = "grey70", size = 0.2, fill = NA),
    axis.text = element_text(margin = margin(t = 4)),
    plot.subtitle = element_textbox(
      margin = margin(t = 4, b = 8)
    )
  )
ggsave(here(base_path, "plots", "lego-colors-area.png"), width = 8, height = 6)

```


