---
title: "Tuskegee Airmen: Locations of Aerial Victories"
output: github_document
---

## Dataset

The locations of the aerial victories of the Tuskegee Airmen during World War 2 have been extracted from documents on [tuskegeeairmen.org](https://tuskegeeairmen.org) and provided as a [tsv file](tuskegee_airmen_victory_locations.tsv).

| Variable | Description |
|----------|-------------|
| date            | Date of the aerial victory        |
| text_location   | location extracted from the text (e.g. "Munich area, Germany") |
| approx_location | approximate location extracted from the text (e.g. "Munich, Germany")  |
| lat_lon         | Latitude and longitude provided by Google Maps API (format: "0.00,0.00")         |
| source       | Source document (see section "Sources" below)         |


### Known issues

The location the aerial victories on 26 April 1945 is unclear/imprecise.


## How to use the data

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


```{r message=FALSE}
library(tidyverse)
```


```{r cache=TRUE}

base_path <- here::here("2022", "06")
tuesdata <- tidytuesdayR::tt_load("2022-02-08")

```


Separate Aerial victory credits column and extract dates:

```{r}
aerial_victories <- tuesdata$airmen %>% 
  select(aerial_victory_credits) %>% 
  separate_rows(aerial_victory_credits, sep = "; ") %>% 
  na.omit() %>% 
  separate(aerial_victory_credits, #
           into = c("aerial_victory_credits", "aerial_victory_date"),
           sep = " on ", remove = TRUE) %>%
  mutate(aerial_victory_credits = str_extract(aerial_victory_credits, "\\d+ (1/2)?") %>% 
           str_replace(" 1/2", ".5") %>% 
           as.numeric(),
         aerial_victory_date = lubridate::mdy(aerial_victory_date)
         ) %>% 
  arrange(aerial_victory_date)

```


Load tsv file with location data and join with aerial victories dataset:

```{r}

locations <- read_tsv(here::here(base_path, "data", "tuskegee_airmen_victory_locations.tsv")) %>% 
  separate(lat_lon, into = c("lat", "lon"), sep = ",") %>% 
  mutate(across(c(lat, lon), as.numeric))


aerial_victories_locations <- aerial_victories %>% 
  inner_join(locations, by = c("aerial_victory_date" = "date")) %>% 
  group_by(approx_location, lat, lon) %>% 
  summarize(aerial_victory_credits = sum(aerial_victory_credits), .groups = "drop")

```


```{r fig.retina=2}

europe <- rnaturalearth::ne_countries(scale = 50, continent = "Europe", 
                                      returnclass = "sf")

ggplot() +
  geom_sf(data = europe, fill = "grey89", size = 0.1) +
  geom_point(data = na.omit(aerial_victories_locations),
             aes(x = lon, y = lat, size = aerial_victory_credits),
             shape = 21, col = "white", fill = alpha("black", 0.9)) +
  coord_sf(xlim = c(5, 35), ylim = c(32, 58)) +
  labs(size = "Aerial victory credits") +
  theme_void()

```



### Sources

<sup>1</sup> https://tuskegeeairmen.org/wp-content/uploads/2020/11/TAI_Resources_112-Victories.pdf<br>
<sup>2</sup> https://tuskegeeairmen.org/wp-content/uploads/2020/11/TAI_Resources_TUSKEGEE-AIRMEN-CHRONOLOGY.pdf